Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-07-09 appendix.tex
update proof of theorem

\begin{table}
\centering
\begin{tabular}{|ll|l|}
\hline
$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau, t)$ & $\Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
\hline
$\emptyset$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{1\}$    &$\emptyset$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{2\}$    &$\emptyset$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{3\}$    &$\emptyset$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
$\{1,2,3\}$&$\emptyset$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
$\{1,2\}$  &$\emptyset$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{1,3\}$  &$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
$\{2,3\}$  &$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
\hline
\end{tabular}
\caption{
The maximal likelihood values given $w=1$ and $y_2=1$.
All values multiplied by $1/32$.}
\label{tab:likelihoods-restricted}
\end{table}

We focus on $x_1$, similar arguments applying to the other variables.
The partial likelihood
\[
\tilde{\ell}_{\tau^*,t^*}(\tau^*, t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau^*, t)
\]
as a function of $x_1$, using Table~\ref{tab:sitepatprob}, is
\[
\tilde{\ell}(x_1) \propto \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1).
\]
By some calculus, the gradient with respect to $x_1$ is
\[
\tilde{\ell}_{x_1}(x_1) = \frac{-x_1+x^*(y^*)^2}{1-x_1^2}
\]
which is clearly negative for all $x_1 \ge x^*$.
%TODO: make more explicit/clear why this means \hat{x}_1 \in [0,x^*]
This same process applied to the remaining variables shows the gradient of the full likelihood is negative for $x_1,x_2 \ge x^*$ and $y_1 \ge y^*$ and positive for $y_2,w \ge y^*$, yielding the constraints
\[
\hat{x}_1,\hat{x}_2 \in [0, x^*], \ \hat{y}_1 \in [0, y^*], \ \hat{y}_2,\hat{w} \in [y^*, 1].
\]
\end{proof}

@@

\begin{proof}
%TODO: make this more explicit/clear
We use the fact that $x_1=x_2=x$ as there exists a unique solution and the likelihood is invariant to swapping these two variables.
Let
\[
a_{\emptyset} = \frac{(2x)(y_1+y_2)}{(1+x^2)(1+y_1y_2)}
\]
with similar definitions for the remaining site splits.
The objective function proportional to $w$ is
\begin{align*}
\ell(w) \propto \log(1+w) + \sum_j p_{\siteSplit_j} \log(1 + a_{\siteSplit_j}w)
\end{align*}
and has derivative with respect to $w$ as
\begin{align*}
\ell'(w) := \frac{d}{dw} \ell(w) = \frac{1}{1+w} + \sum_j p_{\siteSplit_j}a_{\siteSplit_j}\frac{1}{1 + a_{\siteSplit_j}w}.
\end{align*}
Note that, $a_{\emptyset} = -a_{13}$, $a_{2} = -a_{123}$, $p_{2} = p_{123}$, and for $x_1=x_2$
\[
a_{1} = a_{3} = a_{12} = a_{23} = 0
\]
so
\begin{align*}
\ell'(w) = \frac{1}{1+w} + \frac{p_{\emptyset}a_{\emptyset}}{1 + a_{\emptyset}w} - \frac{p_{13}a_{\emptyset}}{1 - a_{\emptyset}w} + \frac{p_{2}a_{2}}{1 + a_{2}w} - \frac{p_{2}a_{2}}{1 - a_{2}w}.
\end{align*}
Using the bounds on $\hat{x}_1,\hat{y}_1,\hat{y}_1$, and $\hat{y}_2$ from Lemma~\ref{lemma:bounds}, we have
\[
0 \le a_{\emptyset} \le \frac{2x^*}{1+(x^*)^2},
\]
\[
0 \le a_{2} \le \frac{2x^*}{1+(x^*)^2}\cdot\frac{1}{1-y^*},
\]
so that $\ell'(w)$ has lower bound
\[
\ell'_{L}(w) = \frac{1}{1+w} + - \frac{p_{13}}{c_x^* - w} - \frac{p_{2}}{c_x^*c_y^* - w}
\]
with
\[
c_x^* = \frac{1+(x^*)^2}{2x^*},
\]
\[
c_y^* = 1-y^*.
\]
The solution $\hat{w}$ solving $\ell'(w) = 0$ is greater than $\hat{w}'$ solving $\ell'_{L}(w) = 0$.
As $c_x^* > 1$, provided $c_x^*c_y^* > 1$, $\ell'_{L}(w)$ stays finite and, when considering $\ell'_{L}(w) \lessgtr 0$, we equivalently consider $f(w) \lessgtr 0$ where $f$ is the quadratic function
\begin{align*}
f(w) &= (w)^2\cdot(1+p_{13}+p_{2}) \\
      &\qquad + w\cdot(-c_x^*-c_x^*c_y^*-p_{13}c_x^*c_y^*+p_{13}-p_{2}c_x^*+p_{2}) \\
      &\qquad + ((c_x^*)^2c_y^*-p_{13}c_x^*c_y^*-p_{2}c_x^*).
\end{align*}
We see
\[
\ell'_L(0) = 1 - \frac{p_{13}}{c_x^*} - \frac{p_{2}}{c_x^*c_y^*} > 0
\]
provided again that $c_x^*c_y^* > 1$.
Now this quadratic has either zero, one, or two solutions.
If there are zero solutions, the derivative with respect to $w$ is increasing on all of $[0,1]$ so that $\hat{w} = 1$ and we are done.
If there are one or two solutions,
if the smaller of the zeros of $f(w)$ is greater than one, then $\hat{w} \equiv 1$.
Using the quadratic formula with
\[
a = 1+p_{13}+p_{2},
\]
\[
b = -c_x^*-c_x^*c_y^*-p_{13}c_x^*c_y^*+p_{13}-p_{2}c_x^*+p_{2},
\]
\[
c = (c_x^*)^2c_y^*-p_{13}c_x^*c_y^*-p_{2}c_x^*,
\]
the smaller zero is
\[
\hat{w} = \frac{-b - \sqrt{b^2 - 4ac}}{2a},
\]
which is a function of the generating parameters $x^*$ and $y^*$.
We see that $a \ge 0$ and, provided that $2a+b \leq 0$, we have,
\[
\hat{w} \ge 1 \iff
|2a + b| \leq \sqrt{b^2 - 4ac} \iff
%-b - \sqrt{b^2 - 4ac} \le 2a,
%0 \ge -b - 2a \le \sqrt{b^2 - 4ac},
%b^2 + 4ab + 4a^2 \le b^2 - 4ac,
%a^2 + ab + ac \le 0,
a + b + c \ge 0.
\]
Therefore, we have three conditions to satisfy for $\hat{w} \equiv 1$:
\[
\mathrm{C}1: c_x^*c_y^* > 1,
\]
\[
\mathrm{C}2: 2-c_x^*c_y^*-c_x^*+p_{13}(3-c_x^*c_y^*)+p_{2}(3-c_x^*) \leq 0,
\]
\[
\mathrm{C}3: 1+(c_x^*)^2c_y^*-c_x^*c_y^*-c_x^*+p_{13}(2-2c_x^*c_y^*)+p_{2}(2-2c_x^*) \ge 0.
\]
The set of $x^*,y^*$ satisfying these conditions as well as those for $\emptyset$ being the most likely ancestral state is given in Figure~\ref{fig:incons-analytic}.
\end{proof}

>>>>

\begin{table}
\centering
\begin{tabular}{|ll|l|}
\hline
$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau, t)$ & $\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
\hline
$\emptyset$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{1\}$    &$\emptyset$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{2\}$    &$\emptyset$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{3\}$    &$\emptyset$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
$\{1,2,3\}$&$\emptyset$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
$\{1,2\}$  &$\emptyset$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{1,3\}$  &$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
$\{2,3\}$  &$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
\hline
\end{tabular}
\caption{
The maximal likelihood values given $\emptyset$ is the most likely ancestral state.
All values multiplied by $1/32$.}
\label{tab:likelihoods-restricted}
\end{table}

We focus on $x_1$, similar arguments applying to the other variables.
The partial likelihood
\[
\tilde{\ell}_{\tau^*,t^*}(\tau^*, t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau^*, t)
\]
as a function of $x_1$, using Table~\ref{tab:sitepatprob}, is
\[
\tilde{\ell}(x_1) \propto \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1).
\]
By some calculus, the partial derivative with respect to $x_1$ is
\[
\tilde{\ell}_{x_1}(x_1) = \frac{-x_1+x^*(y^*)^2}{1-x_1^2}
\]
which is clearly negative for all $x_1 \ge x^*$.
%TODO: make more explicit/clear why this means \hat{x}_1 \in [0,x^*]
This same process applied to the remaining variables shows the gradient of the full likelihood is negative for $x_1,x_2 \ge x^*$ and $y_1 \ge y^*$ and positive for $y_2,w \ge y^*$, yielding the constraints
\[
\hat{x}_1,\hat{x}_2 \in [0, x^*], \ \hat{y}_1 \in [0, y^*], \ \hat{y}_2,\hat{w} \in [y^*, 1].
\]
\end{proof}

@@

\begin{proof}
%TODO: make this more explicit/clear
We use the fact that $x_1=x_2=x$ as there exists a unique solution and the likelihood is invariant to swapping these two variables.
Let
\[
a_{\emptyset} = \frac{(2x)(y_1+y_2)}{(1+x^2)(1+y_1y_2)}
\]
with similar definitions for the remaining site splits.
The objective function proportional to $w$ is
\begin{align*}
\ell(w) \propto \log(1+w) + \sum_j p_{\siteSplit_j} \log(1 + a_{\siteSplit_j}w)
\end{align*}
and has derivative with respect to $w$ as
\begin{align*}
\ell'(w) := \frac{d}{dw} \ell(w) = \frac{1}{1+w} + \sum_j p_{\siteSplit_j}a_{\siteSplit_j}\frac{1}{1 + a_{\siteSplit_j}w}.
\end{align*}
Note that, $a_{\emptyset} = -a_{13}$, $a_{2} = -a_{123}$, $p_{2} = p_{123}$, and for $x_1=x_2$
\[
a_{1} = a_{3} = a_{12} = a_{23} = 0
\]
so
\begin{align*}
\ell'(w) = \frac{1}{1+w} + \frac{p_{\emptyset}a_{\emptyset}}{1 + a_{\emptyset}w} - \frac{p_{13}a_{\emptyset}}{1 - a_{\emptyset}w} + \frac{p_{2}a_{2}}{1 + a_{2}w} - \frac{p_{2}a_{2}}{1 - a_{2}w}.
\end{align*}
Using the bounds on $\hat{x}_1,\hat{y}_1,\hat{y}_1$, and $\hat{y}_2$ from Lemma~\ref{lemma:bounds}, we have
\[
0 \le a_{\emptyset} \le \frac{2x^*}{1+(x^*)^2},
\]
\[
0 \le a_{2} \le \frac{2x^*}{1+(x^*)^2}\cdot\frac{1}{1-y^*},
\]
so that $\ell'(w)$ has lower bound
\[
\ell'_{L}(w) = \frac{1}{1+w} - \frac{p_{13}}{c_x^* - w} - \frac{p_{2}}{c_x^*c_y^* - w}
\]
with
\[
c_x^* = \frac{1+(x^*)^2}{2x^*},
\]
\[
c_y^* = 1-y^*.
\]
Clearly $c_x^* > 1$.
We see
\[
\ell'_L(0) = 1 - \frac{p_{13}}{c_x^*} - \frac{p_{2}}{c_x^*c_y^*} > 0
\]
provided $c_x^*c_y^* > 1$.
Note that $\ell'_{L}(w)$ is strictly decreasing in $w$ on the interval $(0,1)$.
Thus, if $\ell'_{L}(1) > 0$ then $\ell(w)$ is increasing on the entire interval $(0,1)$, including at the boundary $w=1$, implying $\hat{w} \equiv 1$.
So we want $x^*$ and $y^*$ where $c_x^*c_y^* > 1$ and $\ell_L'(1) > 0$.
These two conditions are straightforward, the second one being
\[
\ell'_L(1) = \frac{1}{2} - \frac{p_{13}}{c_x^*-1} - \frac{p_{2}}{c_x^*c_y^*-1} > 0.
\]
The set of $x^*,y^*$ satisfying these conditions as well as those for $\emptyset$ being the most likely ancestral state is given in Figure~\ref{fig:incons-analytic}.
\end{proof}
